---
title: "Self-generating CONSORT diagram"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Self-generating CONSORT diagram}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(consort)
data(dispos.data)
```

The goal of this package is to make it easy to create CONSORT diagrams for the transparent reporting of participant allocation in randomized, controlled clinical trials. This is done by creating a standardized disposition data, and using this data as the source for the creation a standard CONSORT diagram. Human effort by supplying text labels on the node can also be achieved. Below is the illustration of the CONSORT diagram creating process for the two different methods.

# Prepare test data
In a clinical research, we have a participants disposition data. One column is the participants' ID, and the following columns indicating the status of the participants at different stage of the study. One can easily derive the number of participants at different stage by counting the number of participants on-study excluding the participants who are excluded. 

# Description

This function developed to populate consort diagram automatically. But to do so, a population disposition data should be prepared. The following data is prepared for demonstration provided in the package. The data can be loaded with `data(dispos.data)`, the variables in this table is explained below: 

-   `trialno`: Participants ID of the participants
-   `exclusion1, exclusion2`: Exclusion reason before and after induction
-   `induction`: Participants ID of the participants who are included in the induction phase, an extra treatment before randomisation.
-   `exclusion`: Exclusion reason before randomisation, including before and after induction
-   `arm, arm3`: Arms pariticipants randomised to.
-   `sbujid_dosed`: Participants ID of the participants who had at least one dose of the protocol treatment.
-   `subjid_notdosed`: Reason for participants not dosed.
-   `followup`: Participants ID planned for follow-up.
-   `lost_followup`: Reason for participants not dosed.
-   `assessed`: Participants ID participants attended assessment.
-   `no_value`: Reason for participants missing final assessment.
-   `mitt`: Participants ID included in the mITT analysis.

```{r data-head, echo=FALSE}
head(dispos.data)
```

# Usage

Basic logic:

1.  The vertical node are the number of patients in the current node, no dropout reasons of inclusion reasons should be provided.
2.  Side box is only used for the drop outs. It include information about the number of patients excluded and reasons. 
3.  Any participants with exclusion reasons will not be included in the next vertical node box after excluded. So the subject id can be used multiple times to indicate how many patients left in the current node. 
4.  The node labels, for example visit number or phase, can only horizontally align to a vertical main nodes, not an exclusion box.
5.  If more than 2 treatment allocation is present, all the exclusion box after the allocation will be aligned to the right

**At each stage, number of non-missing values will be counted**. Set `kickoff_sidebox=TRUE` to remove the observations counted in the side box in the remaining nodes. 

# Self-generating function

To generate consort diagram with data.frame, one should prepare a disposition `data.frame`.

```{r example, eval=FALSE}
consort_plot(data,
             orders,
             side_box,
             allocation = NULL,
             labels = NULL,
             cex = 0.8,
             text_width = NULL,
             widths = c(0.1, 0.9))
```

-   `data`: Dataset prepared above
-   `orders`: A named vector or a list, names as the variable and values as labels in the box. The order of the diagram will be based on this. Variables listed here, but not included in other parameters will calculate the number of non-missing values. A list can be provided with multiple variables in each list element and these variables will be stack into a single node. The side box should only have 1 variable. 
-   `side_box`: Variable vector, appeared as side box in the diagram. The next box will be count the number of non-missing values, so users need to make sure this is correct. The pariticpants id variable can be used multiple times, since only the number of non-missing is calculated for the vertical box.
-   `allocation`: Name of the grouping/treatment variable (optional), the diagram will split into branches on this variables.
-   `labels`: Named vector, names is the location of the vertical node excluding the side box. The position location should plus 1 after the allocation variables if the allocation is defined.
-   `kickoff_sidebox`: kickoff of the observation from the data with following a side box.
-   `cex`: Multiplier applied to font size, Default is 0.6
-   `text_width`: A positive integer giving the target column for wrapping lines in the output.

# Manual 
Functions are mainly in three categories, main box, side box and label box. Others include building function. These are the functions used by the self generating function. These box functions require the previous node and text label.
-   `add_box`: add main box, no previous nodes should be provided if this is the first node.
-   `add_side_box`: add exclusion box.
-   `add_split`: add allocation box, all nodes will be split into groups. The label text for this node and following nodes should be a vector with a length larger than 1. 
-   `add_label_box`: add visiting or phasing label given a reference node.
-   `build_grid`: you don't need this unless you want the `grob` object.
-   `build_grviz`: you don't need this unless you want the `Graphviz` code.

# Plotting engine
The package supports two plotting engines: `grid` (`grid` package) and `Graphviz`. The default was `grid`, `plot(g)` (assuming `g` is the consort plot object) will use the `grid` to draw the plot. The coordinates are calculated internally, one can simply use `build_grviz(g)` to get a built `grob` object. This means you can do use `patchwork` package to combine different plots. You can use the example below to add title and footnotes. 

```{r patchwork, eval=FALSE}
library(patchwork)

wrap_elements(build_grid(g)) + plot_annotation(
  title = 'Consort diagram',
  subtitle = 'Flow chart of the XX study',
  caption = 'Disclaimer: None of these plots are insightful'
)
```

Although many efforts have been made to draw the plot as easy as possible, you may still find some plots are difficult to draw. You may not be happy about the final output or want the plot in a `svg` format to be used in the websites. You can simply `plot(g, grViz = TRUE)` to draw the plot with `Graphviz`. You need to install [DiagrammeR](https://CRAN.R-project.org/package=DiagrammeR) package to do so. If you want to do some twicking to the plot, you can use `build_grviz(g)` to extract the `dot` file and edit. You can use `Rstudio` to render the plot or other tools you prefer. Use the example below to dump the `dot` file.
```{r dump-dotfile, eval = FALSE}
cat(build_grviz(g), file = "consort.gv")
```

# Working example (self generation)
## Single arm

```{r single-arms, message=FALSE, fig.width  = 6, fig.height = 6}
out <- consort_plot(data = dispos.data,
                    orders = c(trialno = "Population",
                               exclusion = "Excluded",
                               trialno = "Allocated",
                               subjid_notdosed = "Not dosed",
                               followup = "Followup",
                               lost_followup = "Not evaluable for the final analysis",
                               mitt = "Final Analysis"),
                    side_box = c("exclusion", "subjid_notdosed", "lost_followup"),
                    cex = 0.9)
plot(out)
```

## Multiple phase and multiple arms

```{r multiple-phase, fig.width  = 9, fig.height = 6}
g <- consort_plot(data = dispos.data,
                  orders = c(trialno = "Population",
                             exclusion1    = "Excluded",
                             induction   = "Induction",
                             exclusion2    = "Excluded",
                             arm3     = "Randomized patient",
                             subjid_notdosed = "Not dosed",
                             mitt = "Final miTT Analysis"),
                  side_box = c("exclusion1", "exclusion2", "subjid_notdosed"),
                  allocation = "arm3",
                  labels = c("1" = "Screening", "2" = "Month 4",
                             "3" = "Randomization", 
                             "5" = "End of study"),
                  cex = 0.7)
plot(g)
```

## Multiple stratification with node stack
In some cases, two level randomisation/stratification is desired. A typical case is the factorial design. Simply provide two allocation variables, but more than two splits are not supported. In a more general cases, one may want to keep the count inside the vertical box instead of excluding. In this cases, one can provide multiple variables in a list. The first variable will simply be counted, the remaining variables will be itemised. 
```{r multiple-split, fig.width  = 13, fig.height = 10, out.width = "95%"}
# We will exclude one arm to avoid too many arms.
df <- dispos.data[!dispos.data$arm3 %in% "Trt C",]
p <- consort_plot(data = df,
                  orders = list(c(trialno = "Population"),
                                c(exclusion = "Excluded"),
                                c(arm     = "Randomized patient"),
                                # The following two variables will be stacked together
                                c(arm3     = "", # Should not provide a value to show the actual arm
                                  subjid_notdosed="Participants not treated"),
                                # The following two variables will be stacked together
                                c(followup    = "Pariticpants planned for follow-up",
                                  lost_followup = "Reason for tot followed"),
                                c(assessed = "Assessed for final outcome"),
                                c(no_value = "Reason for not assessed"),
                                c(mitt = "Included in the mITT analysis")),
                  side_box = c("exclusion", "no_value"), 
                  allocation = c("arm", "arm3"), # Two level randomisation
                  kickoff_sidebox = FALSE,
                  labels = c("1" = "Screening", "2" = "Randomization",
                             "5" = "Follow-up", "7" = "Final analysis"),
                  cex = 0.7)

plot(p)
```

# Working example (human effort)
The previous is to easily generate a consort diagram based on a disposition data, here we show how to create a consort diagram by providing the label text manually. 

## Provide text
```{r providetext1, fig.width  = 7, fig.height = 4}
library(grid)
# Might want to change some settings
options(txt_gp = gpar(cex = 0.8))

txt0 <- c("Study 1 (n=160)", "Study 2 (n=140)")
txt1 <- "Population (n=300)"
txt1_side <- "Excluded (n=15):\n\u2022 MRI not collected (n=3)\n\u2022 Tissues not collected (n=4)\n\u2022 Other (n=8)"

# supports pipeline operator
g <- add_box(txt = txt0) |>
  add_box(txt = txt1) |>
  add_side_box(txt = txt1_side) |> 
  add_box(txt = "Randomized (n=200)") |> 
  add_split(txt = c("Arm A (n=100)", "Arm B (n=100)")) |> 
  add_side_box(txt = c("Excluded (n=15):\n\u2022 MRI not collected (n=3)\n\u2022 Tissues not collected (n=4)\n\u2022 Other (n=8)",
                       "Excluded (n=7):\n\u2022 MRI not collected (n=3)\n\u2022 Tissues not collected (n=4)")) |> 
  add_box(txt = c("Final analysis (n=85)", "Final analysis (n=93)")) |> 
  add_label_box(txt = c("1" = "Screening",
                        "3" = "Randomized",
                        "4" = "Final analysis"))
plot(g)
```

## Missing nodes and multiple split
There might be some cases that the nodes will be missing, this can be handled as well. You can also have multiple splits, but multiple splits for the `grid` hasn't been implemented yet. You can use `plot(g, grViz = TRUE)` to produce the consort plot.

```{r providetext2}
g <- add_box(txt = c("Study 1 (n=8)", "Study 2 (n=12)", "Study 3 (n=12)"))
g <- add_box(g, txt = "Included All (n=20)")
g <- add_side_box(g, txt = "Excluded (n=7):\n\u2022 MRI not collected (n=3)")
g <- add_box(g, txt = "Randomised")
g <- add_split(g, txt = c("Arm A (n=143)", "Arm B (n=142)"))
g <- add_box(g, txt = c("", "From Arm B"))
g <- add_box(g, txt = "Combine all")
# Length needs to be the same as previous one, use a list here.
g <- add_split(g, txt = list(c("Process 1 (n=140)", "Process 2 (n=140)",
                               "Process 3 (n=142)")))

plot(g, grViz = TRUE)

```

## Two-level stratification
Some studies may first stratify patients then randomise, a factorial design for example. This will require two-level stratification. One can simply provide **a list** of the same length of the previous node after a split. More than two splits are not supported. 
```{r manual-multisplit, fig.width  = 13, fig.height = 6, out.width = "80%"}
g <- add_box(txt = "Patient consented (n=200)")
g <- add_side_box(g, txt = "Excluded (n=10):\n\u2022 MRI not collected (n=3)\n\u2022 Other (n=7)")
g <- add_box(g, txt = "Randomised (n=190)")
g <- add_split(g, txt = c("Randomised study (n=144)", "Preference study (n=146)"))

# Provide a list here
g <- add_split(g, txt = list(c("Allocated to surgery (n=70)", 
                                "Allocated to medicine (n=74)"),
                             c("Allocated to surgery (n=47)", 
                              "Allocated to medicine (n=48)",
                               "Placebo (n=51)")))

g <- add_side_box(g, txt = c("Excluded (n=3):\n\u2022 Withdrawn before surgery (n=2)\n\u2022 Declined (n=1)", 
                             "", 
                             "Excluded (n=1):\n\u2022 Withdrawn before surgery (n=1)",
                             "Excluded (n=3):\n\u2022 Withdrawn before surgery (n=1)\n\u2022 Declined (n=1)",
                             "Excluded (n=10):6\n\u2022 Declined (n=10)"),
                  side = rep("right", 5))
g <- add_box(g, txt = c("Analysed (n=67)", "Analysed (n=74)",
                        "Analysed (n=46)", "Analysed (n=45)", "Analysed (n=41)"))

plot(g)

```

## Using disposition table
```{r disposition-data, fig.width  = 7, fig.height = 6.5}
options(txt_gp = gpar(cex = 0.8))

dispos.data$arm <- factor(dispos.data$arm)

txt <- gen_text(dispos.data$trialno, label = "Patient consented")
g <- add_box(txt = txt)

txt <- gen_text(dispos.data$exclusion, label = "Excluded", bullet = TRUE)
g <- add_side_box(g, txt = txt)   

g <- add_box(g, txt = gen_text(dispos.data$arm, label = "Patients randomised")) 

txt <- gen_text(dispos.data$arm)
g <- add_split(g, txt = txt)

# Exclude subjects
dispos.data <- dispos.data[is.na(dispos.data$exclusion), ]

txt <- gen_text(split(dispos.data[,c("subjid_notdosed")], dispos.data$arm),
                label = "Not dosed", bullet = TRUE)
g <- add_box(g, txt = txt, just = "left")

txt <- gen_text(split(dispos.data$mitt, dispos.data$arm),
                label = "Primary mITT analysis")
g <- add_box(g, txt = txt)

g <- add_label_box(g, txt = c("1" = "Baseline",
                              "5" = "Final analysis"))

plot(g)
```

## For Shiny and HTML
Although all the efforts has been made to precisely calculate the coordinates of the nodes, it is not very accurate due to limit of my own knowledge. But you can utilize the [DiagrammeR](https://CRAN.R-project.org/package=DiagrammeR) to produce plots for Shiny and HTML by setting `grViz = TRUE` in `plot`. You can get `Graphviz` code with `build_grviz` of the plot. In addition, use [DiagrammeRsvg](https://CRAN.R-project.org/package=DiagrammeRsvg) and [rsvg](https://CRAN.R-project.org/package=rsvg) save plot in various formats.
```{r save-plot}
plot(g, grViz = TRUE)
```

## Use with Quarto 
Quarto has native [support](https://quarto.org/docs/authoring/diagrams.html) for embedding Graphviz diagrams. You can plot the flowchart without any printing method. 

````{verbatim, lang = "markdown"}
```{r}
cat(build_grviz(g), file = "consort.gv")
```

```{dot}
//| label: consort-diagram
//| fig-cap: "CONSORT diagram of study XXX"
//| file: consort.gv
```

````

# Saving plot
In order to export the plot to fit a page properly, you need to provide the width and height of the output plot. You might need to try different width and height to get a satisfying plot.You can use R basic device destination for the output. Below is how to save a plot in `png` format:
```{r eval=FALSE}
# save plots
png("consort_diagram.png", width = 29, 
    height = 21, res = 300, units = "cm", type = "cairo") 
plot(g)
dev.off() 

```

Or you can use `ggplot2::ggsave` function to save the plot object:
```{r eval=FALSE}
ggplot2::ggsave("consort_diagram.pdf", plot = build_grid(g))
```

Or save with [DiagrammeRsvg](https://CRAN.R-project.org/package=DiagrammeRsvg) and [rsvg](https://CRAN.R-project.org/package=rsvg) to `png` or `pdf`
```{r eval=FALSE}
plot(g, grViz = TRUE) |> 
    DiagrammeRsvg::export_svg() |> 
    charToRaw() |> 
    rsvg::rsvg_pdf("svg_graph.pdf")
```
