---
title: "Customising CONSORT Diagrams"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Customising CONSORT Diagrams}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(consort)
library(grid)
```

This vignette covers customisation options for CONSORT diagrams created with the `consort` package. For the basics of building diagrams, see `vignette("consort_diagram")`.

# Controlling Diagram Size

By default, the diagram fills 98% of the available device area. You can control the viewport size using the `diagram_width` and `diagram_height` parameters in `plot()` or `build_grid()`.

## Proportional sizing

Pass a number between 0 and 1 to scale relative to the device:

```{r proportional-width, fig.width = 8, fig.height = 5}
g <- add_box(txt = "Assessed for eligibility (n=450)") |>
  add_side_box(txt = "Excluded (n=60):\n\u2022 Not meeting criteria (n=30)\n\u2022 Declined (n=20)\n\u2022 Other (n=10)") |>
  add_box(txt = "Randomized (n=390)") |>
  add_split(txt = c("Allocated to intervention (n=195)", "Allocated to control (n=195)")) |>
  add_box(txt = c("Received intervention (n=190)", "Received control (n=193)"))

# Use 70% of device width, full height
plot(g, diagram_width = 0.7)
```

## Absolute sizing with grid units

For precise output dimensions, pass a `grid::unit()` object. Supported units include `"in"` (inches), `"cm"`, `"mm"`, and `"pt"`:

```{r absolute-width, eval = FALSE}
# Fixed 5 inch wide, 7 inch tall viewport
plot(g, diagram_width = unit(5, "in"), diagram_height = unit(7, "in"))

# Fixed centimetre dimensions
plot(g, diagram_width = unit(12, "cm"), diagram_height = unit(18, "cm"))
```

## Width control with ggsave

When saving with `ggplot2::ggsave`, use `build_grid()` directly:

```{r ggsave-width, eval = FALSE}
grob_obj <- build_grid(g, diagram_width = 0.8)
ggplot2::ggsave("consort.pdf", plot = grob_obj, width = 10, height = 8)
```

# Styling Text and Boxes

## Global text appearance

Use `options(txt_gp = gpar(...))` to set font properties for all boxes:

```{r text-styling, fig.width = 7, fig.height = 4}
# Smaller text with blue colour
options(txt_gp = gpar(cex = 0.7, col = "navy"))

g <- add_box(txt = "Screened (n=500)") |>
  add_side_box(txt = "Excluded (n=50)") |>
  add_box(txt = "Enrolled (n=450)")

plot(g)

# Reset to defaults
options(txt_gp = gpar())
```

## Global box appearance

Use `options(box_gp = gpar(...))` to change box fill, border colour and line width:

```{r box-styling, fig.width = 7, fig.height = 4}
options(
  txt_gp = gpar(cex = 0.8),
  box_gp = gpar(fill = "lightyellow", col = "steelblue", lwd = 2)
)

g <- add_box(txt = "Assessed for eligibility (n=200)") |>
  add_side_box(txt = "Excluded (n=20)") |>
  add_box(txt = "Randomized (n=180)") |>
  add_split(txt = c("Treatment A (n=90)", "Treatment B (n=90)"))

plot(g)

# Reset
options(txt_gp = gpar(), box_gp = gpar(fill = "white"))
```

## Per-box styling

You can pass `gpar` arguments to individual box functions via `txt_gp` and `box_gp`:

```{r per-box-styling, fig.width = 7, fig.height = 5}
options(txt_gp = gpar(cex = 0.8))

g <- add_box(txt = "Total screened (n=500)",
             box_gp = gpar(fill = "#E8F4FD", col = "#2171B5", lwd = 2)) |>
  add_side_box(txt = "Screen failures (n=100)") |>
  add_box(txt = "Randomized (n=400)") |>
  add_split(txt = c("Drug (n=200)", "Placebo (n=200)")) |>
  add_box(txt = c("Completed (n=185)", "Completed (n=190)"))

plot(g)

options(txt_gp = gpar())
```

# Text Indentation

Use the `indent` parameter in `textbox()` (passed through via `...` in `add_box`, `add_side_box`, `add_split`) to add left indentation to the text inside a box. The value is in millimetres by default, or you can pass a `grid::unit()` object.

## Per-box indentation

```{r indent-per-box, fig.width = 7, fig.height = 4}
options(txt_gp = gpar(cex = 0.8))

g <- add_box(txt = "Screened (n=500)") |>
  add_side_box(txt = "Excluded (n=50):\n\u2022 Not eligible (n=30)\n\u2022 Declined (n=20)",
               indent = 5) |>
  add_box(txt = "Enrolled (n=450)")

plot(g)

options(txt_gp = gpar())
```

## Global indentation

Set `options(box_indent = ...)` to apply indentation to all boxes:

```{r indent-global, fig.width = 7, fig.height = 4}
options(txt_gp = gpar(cex = 0.8), box_indent = 4)

g <- add_box(txt = "Population (n=300)") |>
  add_side_box(txt = "Excluded (n=15)") |>
  add_box(txt = "Randomized (n=285)") |>
  add_split(txt = c("Arm A (n=142)", "Arm B (n=143)"))

plot(g)

# Reset
options(txt_gp = gpar(), box_indent = 0)
```

## Using unit objects for indent

For precise control, pass a `grid::unit()` object with any supported unit (`"in"`, `"cm"`, `"mm"`, `"pt"`):

```{r indent-unit, eval = FALSE}
# Indent by 0.2 inches
add_box(txt = "My text", indent = unit(0.2, "in"))

# Indent by 1 centimetre
add_box(txt = "My text", indent = unit(1, "cm"))
```

# Text Wrapping

Long text labels can make boxes too wide. Use `text_width` to wrap text at a specified column width:

```{r text-wrapping, fig.width = 8, fig.height = 5}
options(txt_gp = gpar(cex = 0.8))

long_text <- "Excluded from analysis due to protocol deviation, withdrawal of consent, or missing primary endpoint data"

g <- add_box(txt = "Enrolled participants in the study (n=350)") |>
  add_side_box(txt = paste0(long_text, " (n=30)"), text_width = 40) |>
  add_box(txt = "Included in primary analysis (n=320)")

plot(g)

options(txt_gp = gpar())
```

The `text_width` parameter is also available in `consort_plot()` and applies to all nodes.

# Working with gen_text

The `gen_text()` function generates formatted labels from data. It is used internally by `consort_plot()`, but is also useful on its own for building manual diagrams from data.

## Basic counts

```{r gen-text-basic}
# Simple count
ids <- c("P001", "P002", "P003", "P004", "P005")
gen_text(ids, label = "Enrolled")
```

## Categorical breakdowns

```{r gen-text-factor}
# With bullet points showing breakdown
reasons <- factor(c("Withdrew consent", "Adverse event", "Withdrew consent",
                     "Lost to follow-up", "Adverse event"))
gen_text(reasons, label = "Discontinued", bullet = TRUE)
```

## Without a label (counts only)

```{r gen-text-nolabel}
# Returns individual category counts
gen_text(reasons)
```

## Split by group

When your data is split by treatment arm, `gen_text()` returns one label per group:

```{r gen-text-split}
arms <- factor(c("Drug", "Drug", "Placebo", "Placebo", "Drug"))
ids_by_arm <- split(ids, arms)
gen_text(ids_by_arm, label = "Randomized")
```

## Nested reasons with data frames

For two-level breakdowns (e.g., reason within subgroup), pass a two-column data frame:

```{r gen-text-dataframe}
df <- data.frame(
  category = factor(c("AE", "AE", "Withdrew", "Withdrew", "AE")),
  detail = c("Headache", "Nausea", "Personal", "Travel", "Fatigue")
)
gen_text(df, label = "Discontinued")
```

## Using gen_text in manual diagrams

```{r gen-text-manual, fig.width = 7, fig.height = 5}
options(txt_gp = gpar(cex = 0.8))

data(dispos.data)

# Build labels from data
txt_enrolled <- gen_text(dispos.data$trialno, label = "Screened")
txt_excluded <- gen_text(dispos.data$exclusion, label = "Excluded", bullet = TRUE)
txt_rand <- gen_text(dispos.data$arm, label = "Randomized")
txt_arms <- gen_text(dispos.data$arm)

g <- add_box(txt = txt_enrolled) |>
  add_side_box(txt = txt_excluded) |>
  add_box(txt = txt_rand) |>
  add_split(txt = txt_arms)

plot(g)

options(txt_gp = gpar())
```

# Complete Examples

## Example 1: Observational cohort study

Not all CONSORT-style diagrams are for randomised trials. Here is an observational cohort study with screening, enrolment and follow-up, using custom styling throughout:

```{r example-observational, fig.width = 8, fig.height = 7}
options(
  txt_gp = gpar(cex = 0.8, col = "grey20"),
  box_gp = gpar(fill = "#F0F4F8", col = "#4A90D9", lwd = 1.5)
)

g <- add_box(txt = "Medical records reviewed (n=1,200)") |>
  add_side_box(txt = "Excluded (n=320):\n\u2022 Incomplete records (n=180)\n\u2022 Age < 18 years (n=85)\n\u2022 Duplicate entries (n=55)",
               indent = 3) |>
  add_box(txt = "Eligible participants (n=880)") |>
  add_side_box(txt = "Declined consent (n=130)") |>
  add_box(txt = "Enrolled in cohort (n=750)") |>
  add_side_box(txt = "Lost to follow-up (n=45):\n\u2022 Relocated (n=28)\n\u2022 Unreachable (n=17)",
               indent = 3) |>
  add_box(txt = "Completed 12-month follow-up (n=705)") |>
  add_label_box(txt = c("1" = "Screening",
                         "3" = "Enrolment",
                         "5" = "Follow-up"))

plot(g)

options(txt_gp = gpar(), box_gp = gpar(fill = "white"))
```

## Example 2: Three-arm trial with detailed exclusions

A trial with three treatment arms, showing left-justified side boxes with bullet-point exclusions and indentation:

```{r example-three-arm, fig.width = 10, fig.height = 7}
options(txt_gp = gpar(cex = 0.7))

g <- add_box(txt = "Assessed for eligibility (n=600)") |>
  add_side_box(txt = "Excluded (n=90):\n\u2022 Not meeting inclusion criteria (n=45)\n\u2022 Declined to participate (n=30)\n\u2022 Other reasons (n=15)",
               indent = 4) |>
  add_box(txt = "Randomized (n=510)") |>
  add_split(txt = c("Arm A: High dose (n=170)",
                     "Arm B: Low dose (n=170)",
                     "Arm C: Placebo (n=170)")) |>
  add_side_box(txt = c("Discontinued (n=12):\n\u2022 Adverse event (n=7)\n\u2022 Withdrew consent (n=5)",
                        "Discontinued (n=8):\n\u2022 Adverse event (n=3)\n\u2022 Withdrew consent (n=5)",
                        "Discontinued (n=15):\n\u2022 Adverse event (n=4)\n\u2022 Withdrew consent (n=11)"),
               indent = 3) |>
  add_box(txt = c("Analysed (n=158)", "Analysed (n=162)", "Analysed (n=155)")) |>
  add_label_box(txt = c("1" = "Enrolment",
                         "3" = "Allocation",
                         "5" = "Analysis"))

plot(g)

options(txt_gp = gpar())
```

## Example 3: Compact diagram with narrow width

When publishing in a journal column, use `diagram_width` with smaller font sizes to fit:

```{r example-compact, fig.width = 6, fig.height = 5}
options(txt_gp = gpar(cex = 0.65))

g <- add_box(txt = "Screened (n=250)") |>
  add_side_box(txt = "Excluded (n=50)") |>
  add_box(txt = "Randomized (n=200)") |>
  add_split(txt = c("Treatment (n=100)", "Control (n=100)")) |>
  add_side_box(txt = c("Lost (n=5)", "Lost (n=8)")) |>
  add_box(txt = c("Analysed (n=95)", "Analysed (n=92)"))

plot(g, diagram_width = 0.6)

options(txt_gp = gpar())
```

## Example 4: Using data with gen_text and custom formatting

Build a complete diagram from the demo dataset, combining `gen_text()` for automatic label generation with manual styling:

```{r example-data-driven, fig.width = 9, fig.height = 7}
options(
  txt_gp = gpar(cex = 0.75),
  box_gp = gpar(fill = "#FFFDF5", col = "grey40")
)

data(dispos.data)

# Build all text labels from the data
txt1 <- gen_text(dispos.data$trialno, label = "Patients screened")
txt2 <- gen_text(dispos.data$exclusion, label = "Excluded", bullet = TRUE)
txt3 <- gen_text(dispos.data$arm, label = "Randomized")
txt4 <- gen_text(dispos.data$arm)

# Subset for per-arm labels
dispos_clean <- dispos.data[is.na(dispos.data$exclusion), ]
txt5 <- gen_text(split(dispos_clean$subjid_notdosed, dispos_clean$arm),
                 label = "Not dosed", bullet = TRUE)
txt6 <- gen_text(split(dispos_clean$mitt, dispos_clean$arm),
                 label = "mITT population")

g <- add_box(txt = txt1) |>
  add_side_box(txt = txt2, indent = 3) |>
  add_box(txt = txt3) |>
  add_split(txt = txt4) |>
  add_side_box(txt = txt5, indent = 3) |>
  add_box(txt = txt6) |>
  add_label_box(txt = c("1" = "Screening",
                         "3" = "Randomization",
                         "5" = "Treatment",
                         "6" = "Analysis"))

plot(g)

options(txt_gp = gpar(), box_gp = gpar(fill = "white"))
```

## Example 5: Graphviz output for web and Shiny

The same diagram can be rendered with Graphviz for interactive web use. This produces cleaner SVG output and supports Shiny:

```{r example-grviz}
g <- add_box(txt = "Assessed (n=400)") |>
  add_side_box(txt = "Excluded (n=50)") |>
  add_box(txt = "Randomized (n=350)") |>
  add_split(txt = c("Drug A (n=175)", "Drug B (n=175)")) |>
  add_side_box(txt = c("Withdrew (n=10)", "Withdrew (n=15)")) |>
  add_box(txt = c("Completed (n=165)", "Completed (n=160)"))

plot(g, grViz = TRUE)
```

## Example 6: Combining features — indent, wrapping and styling

This example shows how indent, text wrapping and per-box styling can be combined in one diagram:

```{r example-combined, fig.width = 8, fig.height = 7}
options(txt_gp = gpar(cex = 0.75))

g <- add_box(txt = "Patients identified from electronic health records across 12 participating sites (n=2,500)",
             text_width = 45,
             box_gp = gpar(fill = "#E3F2FD", col = "#1565C0", lwd = 2)) |>
  add_side_box(txt = "Excluded (n=600):\n\u2022 Did not meet age criteria 18-75 years (n=220)\n\u2022 Prior enrolment in competing trial (n=185)\n\u2022 Unable to provide informed consent (n=110)\n\u2022 Other protocol exclusions (n=85)",
               text_width = 45,
               indent = 4) |>
  add_box(txt = "Eligible and consented (n=1,900)") |>
  add_split(txt = c("Intervention (n=950)", "Standard care (n=950)")) |>
  add_side_box(txt = c("Discontinued (n=60):\n\u2022 Adverse events (n=25)\n\u2022 Withdrew consent (n=20)\n\u2022 Protocol deviation (n=15)",
                        "Discontinued (n=45):\n\u2022 Adverse events (n=15)\n\u2022 Withdrew consent (n=18)\n\u2022 Lost to follow-up (n=12)"),
               text_width = 38,
               indent = 3) |>
  add_box(txt = c("Included in ITT analysis (n=950)",
                   "Included in ITT analysis (n=950)")) |>
  add_label_box(txt = c("1" = "Identification",
                         "3" = "Randomization",
                         "5" = "Analysis"))

plot(g, diagram_width = 0.9)

options(txt_gp = gpar())
```

# Combining with patchwork

The `build_grid()` function returns a grob that can be combined with other plots using the `patchwork` package:

```{r patchwork-example, eval = FALSE}
library(patchwork)

g <- add_box(txt = "Screened (n=500)") |>
  add_side_box(txt = "Excluded (n=50)") |>
  add_box(txt = "Enrolled (n=450)")

wrap_elements(build_grid(g)) + plot_annotation(
  title = "CONSORT Flow Diagram",
  subtitle = "Randomized controlled trial of Treatment X vs Placebo",
  caption = "Figure 1"
)
```

# Saving Diagrams

## Using base R graphics device

```{r save-base, eval = FALSE}
png("consort.png", width = 29, height = 21, res = 300, units = "cm", type = "cairo")
plot(g)
dev.off()

pdf("consort.pdf", width = 10, height = 8)
plot(g)
dev.off()
```

## Using ggsave with width control

```{r save-ggsave, eval = FALSE}
# Control both the output file dimensions and the diagram viewport
grob_obj <- build_grid(g, diagram_width = 0.85, diagram_height = 0.9)
ggplot2::ggsave("consort.pdf", plot = grob_obj, width = 10, height = 8)
```

## Saving Graphviz output as SVG or PDF

```{r save-grviz, eval = FALSE}
plot(g, grViz = TRUE) |>
  DiagrammeRsvg::export_svg() |>
  charToRaw() |>
  rsvg::rsvg_pdf("consort_grviz.pdf")
```

# Summary of Key Parameters

| Parameter | Where | What it controls |
|-----------|-------|------------------|
| `diagram_width` | `plot()`, `build_grid()` | Overall diagram viewport width (0-1 or `unit()`) |
| `diagram_height` | `plot()`, `build_grid()` | Overall diagram viewport height (0-1 or `unit()`) |
| `text_width` | `add_box()`, `add_split()`, `add_side_box()`, `consort_plot()` | Character column for text wrapping within boxes |
| `txt_gp` | `textbox()`, `options()` | Text appearance (`gpar`: font size, colour, face) |
| `box_gp` | `textbox()`, `options()` | Box appearance (`gpar`: fill, border, line width) |
| `cex` | `consort_plot()` | Font size multiplier (shorthand for `txt_gp`) |
| `indent` | `textbox()`, `options()` | Left indentation in mm or `unit()` |
| `just` | `add_box()` | Text justification: `"center"`, `"left"`, `"right"` |
